---
title: "run_analysis"
author: "S. Kiran Koya"
date: "October 22, 2015"
output: 
  html_document: 
    highlight: pygments
    theme: readable
---

run_analysis.R function will perform the following tasks. 

1. Download data from the url
2. Merges the training and the test sets to create one data set.
3. Extracts only the measurements on the mean and standard deviation for each measurement. 
4. Uses descriptive activity names to name the activities in the data set
5. Appropriately labels the data set with descriptive variable names. 
6. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

#Clean up environment and load libraries

```{r}
rm(list = ls(all = TRUE))
library(plyr) 
library(data.table) 
library(dplyr)
```

#1. Download data from the url

```{r}
#Check to see if file is downloaded. If not download the data from the url  
filename <- "UCI HAR Dataset"
zfilename <- "getdata-projectfiles-UCI HAR Dataset.zip"
if(! file.exists(zfilename)){
        message("File doesn't exist. Downloading compressed file from the archive")
        fileURL = "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
        download.file(url = fileURL, destfile = zfilename)
}

#Check to see if file is uncompressed. If not extract data. 
if(! file.exists(filename)){
        message("Extracting the data")
        unzip(zipfile = zfilename)
}
rm(filename, zfilename)

#read feature and activity file
features <- read.table("UCI HAR Dataset/features.txt")
activity <- read.table("UCI HAR Dataset/activity_labels.txt")

#read test files
X_test <- read.table("UCI HAR Dataset/test/X_test.txt")
Y_test <- read.table("UCI HAR Dataset/test/Y_test.txt")
subject_test <- read.table("UCI HAR Dataset/test/subject_test.txt")

#read train files
X_train <- read.table("UCI HAR Dataset/train/X_train.txt")
Y_train <- read.table("UCI HAR Dataset/train/Y_train.txt")
subject_train <- read.table("UCI HAR Dataset/train/subject_train.txt")
```

#2. Merge the training and the test sets to create one data set.
```{r}
X_all <- rbind(X_train, X_test) #merge both X files
Y_all <- rbind(Y_train, Y_test) #merge both Y files
subjects_all <- rbind(subject_test, subject_train) #merge both Subject files
rm(X_test,Y_test,X_train,Y_train,subject_train,subject_test) #remove unnecessary files
```

#3. Appropriately label the data set with descriptive variable names.
```{r}
names(X_all) = features[,2] #replace variable names with corresponding feature names
names(Y_all) = c("activityID") #name variable as activityID
names(subjects_all) = c("subjects") #name variable as subjects
names(activity) = c("activityID", "activityName") #name variables as "activityID" and "activityName"
```

#4. Extract only the mean and standard deviation for each measurement. 
```{r}
mindex <- grep("mean",features[, 2]) #get any feature that has mean in it 
mindex <- append(mindex, 555:561) #555:561 are also means but were not identified by above command. Add them to the index. 
sindex <- grep("std",features[, 2]) #get any feature that has std in it. 
mergedindex<- unique(c(mindex,sindex)) # Remove duplicates and merge indeces.
#using the mergedindex extract column data for means and stds. 
X_all_diced <- X_all[mergedindex]
rm(mindex, sindex, mergedindex) #remove unnecessary files
```

#5. Use descriptive activity names to name the activities in the data set
```{r}
#From Y_all_labeled table, take the activity and merge it with X_all_diced table to the corresponding activity. 
Y_all_labeled <- merge(Y_all, activity, "activityID") #Join Y_all table with activity table by "activityID"
X_all_diced$Activity <- Y_all_labeled[,2] #Create a new column Activity and assign corresponding activity

#From Subjects_all table, take subjects and appropritate them to X_all_diced table subjects. 
X_all_diced$Subjects <- subjects_all[, 1] #create a new variable Subjects and add corresponding description
X_all_diced$Subjects <- sub("^", "Participant ", X_all_diced$Subjects) #change subject names to descriptive "Participant 1 etc."
rm(X_all, Y_all, Y_all_labeled, subjects_all, activity, features) #remove unnecessary files.
```
###clean up names and descriptive terms. 
```{r}
#clean up variable names to make them more descriptive using gsub function
names(X_all_diced) <- gsub("Acc", "_Accelerator", names(X_all_diced)) 
names(X_all_diced) <- gsub("Mag", "_Magnitude", names(X_all_diced))
names(X_all_diced) <- gsub("Gyro", "_Gyroscope", names(X_all_diced))
names(X_all_diced) <- gsub("^t", "Time_", names(X_all_diced))
names(X_all_diced) <- gsub("^f", "Frequency_", names(X_all_diced))
names(X_all_diced) <- gsub("std", "StDev", names(X_all_diced))
names(X_all_diced) <- gsub("mean", "Mean", names(X_all_diced))
```

#6. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.
```{r}
X_all_diced.dt <- data.table(X_all_diced)
TidyData <- X_all_diced.dt[, lapply(.SD, mean), by = 'Subjects,Activity']
write.table(TidyData, file = "Processed_Data.txt", row.names = FALSE)
rm(X_all_diced, X_all_diced.dt) #remove unnecessary files. 
names(TidyData)
```




